Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-10-26 paper/main.tex
record transformer constraints

\begin{definition}\cite{nlab:action, grimaldiDiscreteCombinatorialMathematics2006}
  An \textcolor{action}{\textbf{action}} \footnote{Throughout this paper, we augment the math with color to group conceptually similar objects and functions\cite{headMathAugmentationHow2022}, for example \textcolor{action}{actions} and \textcolor{action}{sets of actions}. This color coding carries through to the figures.} of \textcolor{action}{$G$} on $X$ is a function  $act: \textcolor{action}{G} \times X \rightarrow X$. An action has the properties of identity $act(\textcolor{action}{e}, x) = x$ for all  $x \in X$ and associativity $act(\textcolor{action}{g}, act(\textcolor{action}{f}, x)) = act(\textcolor{action}{f} \circ \textcolor{action}{g}, x)$ for $\textcolor{action}{f},\textcolor{action}{g} \in \textcolor{action}{G}$.
\end{definition}

@@

To build on the rich typing system provided by Spivak, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the set of all data records
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. 
Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accesed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a topological space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.

@@

Each section of a sheaf over a point returns a single record in the fiber. The sheaf over an open set $\openset$ surrounding a point $\dbasepoint$ is called a \textit{stalk}\cite{StalkSheaf2019,harder2008lectures}
\begin{equation}
  \label{eq:atct:sheaf:stalk}
    \sheaf_{\dbase, \dtotalc}\restriction_{\dbasepoint}\coloneqq \lim\limits_{\openset\ni \dbasepoint} \Gamma(\openset, \dtotal\restriction_{\openset}) 
\end{equation}
where the fiber is contained inside the stalk  $\dfiber_{\dbasepoint} \subset  \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$. The \textit{germ} is the section evaluated at a point in the stalk  $\dsection(\dbasepoint) \in \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$ and is the data. Since the stalk and the germ include the values near the limit of the point at \dbasepoint, the germ can be used to compute the mathematical derivative of the data for visualization tasks that require this information.

@@

Functors between sheaves are a way of expressing the bookkeeping involved in keeping track of which graphic section \gsection\ corresponds to which data section \dsection. The $(\dbasepoint_i, \gbase_i)$ pairing expressed in \autoref{eq:atct:xi} establishes that there is a correspondence between sections evaluated over $\dbasepoint_i$ and $\gbase_i$. This allows us to construct graphic specifications for each data index $\vindexpush\gsection$ and retrieve the data $\vindexpull\dsection$ for any graphic section generating any piece of a graphic. In \autoref{fig:atct:morphisms:sheaf}, the visualization is a graphic representation of a unit circle and the sin and cosine curves on that interval. The index lookup $\vindex$ describes which parts of the circle and curves are generated from which points on the unit circle. Given this correspondance, the pullback $\vindexpull$ looks up which values are being represented in a given part of the graphic. The pushforward $\vindexpush$ describes how a graphic is supposed to look for each point in the data space. The graphic parameterized in \autoref{fig:atct:morphisms:sheaf} is intended as an approximation of $\vindexpush\gsection$.

Visualization specification languages such as vega \cite{heerDeclarative2010} and svg \cite{quintScalable2003} assume that there is a graphic  specification for every point. Interactive tooltips assume that there is a map between the graphics rendered on screen and the data such that a glyph maps back to a record. The expectation that these mappings between data and graphic space exist are encapsulated in the functors $(\vindex, \vindexpush, \vindexpull)$.

@@

\paragraph{\textbf{Topological structure}}
\noindent
The base space transformation is a pointwise continuous map from one open set to another open set in the same base space
\begin{equation}
  \label{eq:atct:morphism:base}
 \dfunch: \openset^{\prime}\rightarrow \openset
 \end{equation}
such that $\openset, \openset^{\prime} \subseteq \dbase$. This means $\openset$ and $\openset^{\prime}$ are of the same topology type. To correctly align the sections with the remapped base space, there is a a corresponding section pullback function
\begin{equation}
  \label{eq:atct:morphism:basepull}
  \dfuncpull \dsection \restriction_{\openset^{\prime}}: \dsection\restriction_{\openset^{\prime}} \mapsto \dsection \restriction_{\openset^{\prime} \circ \dfunch} 
\end{equation}
such that $\dsection|_{\openset} = \dfuncpull\dsection|_{\opensetc^{\prime}}$ because $\dsection|_{\openset} = \dsection|_{\dfunch(\opensetc^{\prime})}$. This means that the base space transformation \dfunch\ does not change the data values at a given point
\begin{equation}
  \label{eq:atct:morphism:verify_base}
  \dsection(\dbase) = \dfuncpull\dsection(\dbasepoint^{\prime}) = \dsection(\dfunch(\dbasepoint^{\prime}))
\end{equation} 
where $\dfunch(\dbasepoint^{\prime}) = \dfunch(\dbasepoint)$.

\paragraph{\textbf{Records}}
\noindent
As introduced in \autoref{eq:artist:sheaves:monoid_morphism}, the fiber transformation $\dfunct: \dfuncpull \dtotal_{\dbasepoint^{\prime}} \rightarrow \dfuncpull \dtotal_{\dbasepoint^{\prime}}$ is a monoid action on the fiber $\dfunct \in Hom(\dfuncpull\dfiber\restriction_{\dbasepoint^{\prime}},\dfuncpull\dfiber\restriction_{\dbasepoint^{\prime}})$ restricted to a point $\dbasepoint^{\prime} \in \openset^{\prime}$. The fiber transformation is a change in section 
\begin{equation}
  \label{eq:atct:morphism:fiber}
  \dfunct: \dfuncpull \dsection \restriction_{\openset^{\prime}} \mapsto \dfuncpull \dsection^{\prime} \restriction_{\openset}
\end{equation}
where $\dsection, \dsection^{\prime} \in \Gamma(\openset^{\prime}, \dfuncpull\dtotal\restriction_{\openset^{\prime}})$. Since there are many different fiber transformations,  there is not a uniform way of expressing how this morphism acts on sections unlike in \autoref{eq:atct:morphism:verify_base}. Examples of a fiber transformation are actions such as partial ordering for ranking purpose \cite{bruggemannRankingPrioritizationMultiindicator2011} and the permutation, ordering, translation, and scaling codified as Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}.

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | c c c|}
    \hline
    &\textcolor{base}{point}/\textcolor{base}{openset}/\textcolor{base}{base space} & \textcolor{fiber}{fiber space} & \textcolor{total}{total space}\\
     &  location/subset/indicies & fields &  dataset type\\
    \hline
   Data & $\dbasepointc \in \opensetc \subseteq \dbasec$ & \dfiberc & \dtotalc\\
   Visual & $\dbasepointc \in \opensetc \subseteq \dbasec$  & \vfiberc & \vtotalc\\
   Graphic & $\gbasepointc \in \opensetgc \subseteq \gbasec$ & \gfiberc & \gtotalc\\
   \hline
  \end{tabular}
  \caption{Topological spaces introduced in \autoref{sec:atct:fiber-bundles}}
  \label{tab:appendix:summary:objects}
  }
\end{table}

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l  l |}
\hline
& function & constraint\\
\hline
 \gbasepointc\ to \dbasepointc& $ \vindexc: \opensetgc \rightarrow \opensetc$ & for $\dbasepointc \in \opensetc$ exists $\gbasepointc \in \opensetgc$ \\
 & & s.t. $\vindexc(\gbasepointc) = \dbasepointc$\\
 graphic for \dbasepointc & $ \vindexpushc \gsectionc: \opensetc \rightarrow \vindexpushc\gtotalc\restriction_{\opensetc}$ & $\vindexpushc\gsectionc(\dbasepointc)(\gbasepointc) = \gsectionc(\gbasepointc)$ \\
 record for \gbasepointc & $\vindexpullc\dsectionc: \opensetgc \rightarrow  \vindexpullc\dtotalc\restriction_{\opensetgc}$ & $\vindexpullc\dsectionc(\gbasepointc)=\dsectionc(\vindexc(\gbasepointc)) = \dsectionc(\dbasepointc)$  \\
\hline
  \end{tabular}
  \caption{Functors between graphic and data base spaces  \autoref{sec:atct:xi}}
  \label{tab:appendix:summary:transport}
  }
\end{table}

\begin{table}[h!]
\centering
{\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|}

\end{tabular}
\note{changing locations, moving records to new locations, changing records}
\caption{Functions $\dfunc=(\dfunchc, \dfunctc)$ for modifying data records. Equivalent constructions can be applied to elements in visual and graphic sheaves, and these functions are distinguised through subscripts $\dfunc_{\dtotal}$, $\dfunc_{\vtotal}$ and $\dfunc_{\gtotal}$}
\label{tab:appendix:summary:datamod}
}
\end{table}

>>>>

\begin{definition}\label{def:related-work:action}\cite{nlab:action, grimaldiDiscreteCombinatorialMathematics2006}
  An \textcolor{action}{\textbf{action}} \footnote{Throughout this paper, we augment the math with color to group conceptually similar objects and functions\cite{headMathAugmentationHow2022}, for example \textcolor{action}{actions} and \textcolor{action}{sets of actions}. This color coding carries through to the figures.} of \textcolor{action}{$G$} on $X$ is a function  $act: \textcolor{action}{G} \times X \rightarrow X$. An action has the properties of identity $act(\textcolor{action}{e}, x) = x$ for all  $x \in X$ and associativity $act(\textcolor{action}{g}, act(\textcolor{action}{f}, x)) = act(\textcolor{action}{f} \circ \textcolor{action}{g}, x)$ for $\textcolor{action}{f},\textcolor{action}{g} \in \textcolor{action}{G}$.
\end{definition}

@@

To build on the rich typing system provided by Spivak, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the space of all possible data records
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. 
Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accesed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a topological space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.

@@

Each section of a sheaf over a point returns a single record in the fiber. The sheaf over an open set $\openset$ surrounding a point $\dbasepoint$ is called a \textit{stalk}\cite{StalkSheaf2019, harder2008lectures}
\begin{equation}
  \label{eq:atct:sheaf:stalk}
    \sheaf_{\dbase, \dtotalc}\restriction_{\dbasepoint}\coloneqq \lim\limits_{\openset\ni \dbasepoint} \Gamma(\openset, \dtotal\restriction_{\openset}) 
\end{equation}
where the fiber is contained inside the stalk  $\dfiber_{\dbasepoint} \subset  \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$. The \textit{germ} is the section evaluated at a point in the stalk  $\dsection(\dbasepoint) \in \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$ and is the data. Since the stalk and the germ include the values near the limit of the point at \dbasepoint, the germ can be used to compute the mathematical derivative of the data for visualization tasks that require this information.

@@

Functors between sheaves are a way of expressing the bookkeeping involved in keeping track of which graphic section \gsection\ corresponds to which data section \dsection. The $(\dbasepoint_i, \gbase_i)$ pairing expressed in \autoref{eq:atct:xi} establishes that there is a correspondence between sections evaluated over $\dbasepoint_i$ and $\gbase_i$. This allows us to construct graphic specifications for each data index $\vindexpush\gsection$ and retrieve the data $\vindexpull\dsection$ for any graphic section generating any piece of a graphic. In \autoref{fig:atct:morphisms:sheaf}, the visualization is a graphic representation of a unit circle and the sin and cosine curves on that interval. The index lookup $\vindex$ describes which parts of the circle and curves are generated from which points on the unit circle. Given this correspondance, the pullback $\vindexpull\dsection$ looks up which values are being represented in a given part of the graphic. This type of lookup is critical for interactive techniques such as brushing, linking, and tooltips\cite{beckerBrushingScatterplots1987}. The pushforward $\vindexpush\gsection$ describes how a graphic is supposed to look for each point in the data space. The graphic parameterization in \autoref{fig:atct:morphisms:sheaf} is intended as an approximation of $\vindexpush\gsection$ and is akin to declarative visualization specs such as vega \cite{satyanarayanDeclarativeInteractionDesign2014} and svg \cite{quintScalable2003}. These specs and $\vindexpushc\gsection$ provide a renderer independent way of describing the graphic and are therefore useful for standardizing internal representation of the graphic and serializing the graphic for portability.

@@

\paragraph{\textbf{Topological structure}}
\noindent
The base space transformation is a pointwise continuous map from one open set to another open set in the same base space
\begin{equation}
  \label{eq:atct:morphism:base}
 \dfunchc: \dbasepointc^{\prime}\mapsto \dbasepointc
 \end{equation}
such that $\openset, \openset^{\prime} \subseteq \dbase$. This means $\openset$ and $\openset^{\prime}$ are of the same topology type. To correctly align the sections with the remapped base space, there is a a corresponding section pullback function
\begin{equation}
  \label{eq:atct:morphism:basepull}
  \dfuncpullc \dsectionc \restriction_{\opensetc^{\prime}}: \dsectionc\restriction_{\opensetc^{\prime}} \mapsto \dsectionc \restriction_{\opensetc^{\prime} \circ \dfunchc} 
\end{equation}
such that $\dsection|_{\openset} = \dfuncpull\dsection|_{\opensetc^{\prime}}$ because $\dsection|_{\openset} = \dsection|_{\dfunch(\opensetc^{\prime})}$. This means that the base space transformation \dfunch\ does not change the data values at a given point
\begin{equation}
  \label{eq:atct:morphism:verify_base}
  \dsection(\dbase) = \dfuncpull\dsection(\dbasepoint^{\prime}) = \dsection(\dfunch(\dbasepoint^{\prime}))
\end{equation} 
where $\dfunch(\dbasepoint^{\prime}) = \dfunch(\dbasepoint)$.

\paragraph{\textbf{Records}}
\noindent
As introduced in \autoref{eq:artist:sheaves:monoid_morphism}, the fiber transformation $\dfunct: \dfuncpull \dtotal_{\dbasepoint^{\prime}} \rightarrow \dfuncpull \dtotal_{\dbasepoint^{\prime}}$ is a monoid action on the fiber $\dfunct \in Hom(\dfuncpull\dfiber\restriction_{\dbasepoint^{\prime}},\dfuncpull\dfiber\restriction_{\dbasepoint^{\prime}})$ restricted to a point $\dbasepoint^{\prime} \in \openset^{\prime}$. The fiber transformation is a change in section 
\begin{equation}
  \label{eq:atct:morphism:fiber}
  \dfunct: \dfuncpull \dsection \restriction_{\openset^{\prime}} \mapsto \dfuncpull \dsection^{\prime} \restriction_{\openset}
\end{equation}
where $\dsection, \dsection^{\prime} \in \Gamma(\openset^{\prime}, \dfuncpull\dtotal\restriction_{\openset^{\prime}})$. Since there are many different fiber transformations,  there is not a uniform way of expressing how this morphism acts on sections unlike in \autoref{eq:atct:morphism:verify_base}.\note{This is where we reintroduce the equivariance and homomorphism definitions from the introduction}

Examples of a fiber transformation are actions such as partial ordering for ranking purpose \cite{bruggemannRankingPrioritizationMultiindicator2011} and the permutation, ordering, translation, and scaling codified as Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}.

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | c c c|}
    \hline
    &\textcolor{base}{point}/\textcolor{base}{openset}/\textcolor{base}{base space} & \textcolor{fiber}{fiber space} & \textcolor{total}{total space}\\
     &  location/subset/indicies & record/fields &  dataset type\\
    \hline
   Data & $\dbasepointc \in \opensetc \subseteq \dbasec$ & $\delementc \in \dfiberc$ & \dtotalc\\
   Visual & $\dbasepointc \in \opensetc \subseteq \dbasec$  & $\velementc \in \vfiberc$ & \vtotalc\\
   Graphic & $\gbasepointc \in \opensetgc \subseteq \gbasec$ & $\gelementc \in \gfiberc$ & \gtotalc\\
   \hline
  \end{tabular}
  \caption{Topological spaces introduced in \autoref{sec:atct:fiber-bundles}}
  \label{tab:appendix:summary:objects}
  }
\end{table}

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l  l |}
\hline
& function & constraint\\
\hline
 \gbasepointc\ to \dbasepointc& $ \vindexc: \opensetgc \rightarrow \opensetc$ & for $\dbasepointc \in \opensetc$ exists $\gbasepointc \in \opensetgc$ \\
 & & s.t. $\vindexc(\gbasepointc) = \dbasepointc$\\
 graphic for \dbasepointc & $ \vindexpushc \gsectionc: \opensetc \rightarrow \vindexpushc\gtotalc\restriction_{\opensetc}$ & $\vindexpushc\gsectionc(\dbasepointc)(\gbasepointc) = \gsectionc(\gbasepointc)$ \\
 record for \gbasepointc & $\vindexpullc\dsectionc: \opensetgc \rightarrow  \vindexpullc\dtotalc\restriction_{\opensetgc}$ & $\vindexpullc\dsectionc(\gbasepointc)=\dsectionc(\vindexc(\gbasepointc)) = \dsectionc(\dbasepointc)$  \\
\hline
  \end{tabular}
  \caption{Functors between graphic and data indexing spaces  \autoref{sec:atct:xi}}
  \label{tab:appendix:summary:transport}
  }
\end{table}

\begin{table}[h!]
\centering
{\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|r|l|l|}
  \hline
  changes & function & constraints, for all $\dbasepointc \in \opensetc$ \\
  \hline
  index & $\dfunchc: \opensetc \rightarrow \opensetc^{\prime}$ &   $\dsectionc(\dbasepointc) = \dsectionc(\dfunchc(\dbasepointc^{\prime})) = \dfuncpullc \dsectionc(\dbasepointc^{\prime})$ \\
  & & \\
  record  & $\dfunctc: \cgamma{\opensetc^{\prime}}{\dfuncpullc\dtotalc\restriction_{\opensetc}}$ &  $\lim\limits_{x\rightarrow \dbasepointc}\dfunctc(\dsectionc(x)) = \dfunctc(\dsectionc(\dbasepointc))$
  \\
  & $\textcolor{white}{\dfunct: } \rightarrow \cgamma{\opensetc^{\prime}}{\dfuncpullc\dtotalc\restriction_{\opensetc}}$ &    
  \\
  & $\dfunctc: \dfiberc \rightarrow \dfiberc$ &  $\dfunctc(\dsectionc(\dbasepointc)) \in \dfiberc$  \\ 
  & &  $\dfunctc(id_{\dfiberc}(\dsectionc(\dbasepointc))) = id_{\dfiberc}(\dfunctc(\dsectionc(\dbasepointc)))$\\
  & & $\dfunctc(\dfunctc(\dsectionc(\dbasepointc))) = (\dfunctc\circ\dfunctc)(\dsectionc(\dbasepointc))$\\
  \hline
\end{tabular}
\caption{Functions $\dfunc=(\dfunchc, \dfunctc)$ for modifying data records. Equivalent constructions can be applied to elements in visual and graphic sheaves, and these functions are distinguised through subscripts $\dfunc_{\dtotal}$, $\dfunc_{\vtotal}$ and $\dfunc_{\gtotal}$}
\label{tab:appendix:summary:datamod}
}
\end{table}

\begin{table}[h!]
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|lll|}\hline
      scale & property & scale specific constraints \\ \hline
      nominal & permutation &  $\dsectionc(\dbasepointc_1) \neq \dsectionc(\dbasepointc_2) \implies \dfunctc (\dsectionc(\dbasepointc_1)) \neq\dfunctc(\dsectionc(\dbasepointc_2)$)\\
      ordinal &  monotonic &  $\dsectionc(\dbasepointc_1) \leq \dsectionc(\dbasepointc_2) \implies \dfunctc (\dsectionc(\dbasepointc_1)) \leq \dfunctc(\dsectionc(\dbasepointc_2)$) \\
      interval &  translation &  $\dfunctc(\dsectionc(\dbasepointc) + C) = \dfunctc(\dsectionc(\dbasepointc)) + C$ \\
      ratio &  scaling &  $\dfunctc(\dsectionc(\dbasepointc) * C) = \dfunctc(\dsectionc(\dbasepointc))*C $\\ \hline
  \end{tabular}
  \caption{The record transformer $\dfunctc$ is defined by the general constraints listed in \autoref{tab:appendix:summary:datamod} and it must respect any properties that are specific to the record field. For example, this table lists the properties of the Steven's measurement scale types\cite{stevensTheoryScalesMeasurement1946}}
  \label{tab:appendix:summary:datatypes}
\end{table}
